---
editor_options:
  markdown:
    wrap: 72
bibliography: [P:/library-full.bib]     
---

<!-- rmarkdown::render_site(output_format = 'bookdown::pdf_book', encoding = 'UTF-8') -->

# Overview

The literature reviewed above contributes to the research question in three ways. First, the perceptions of reference groups offer a synthesis of how structural positions and social relations with other members of society impact views on social inequality. In this sense, changes in social structure are accompanied by modifications in our social circles, which nurture perceptions of society at large and how individuals perceive their place in the social hierarchy. Second, the structural perspective of social networks offers an analytical approach to the processes that structure sociability, friendship, and kinship networks as well as the behaviour and attitudes of individuals toward groups that ultimately bring societies together. In this sense, structural analysis based on social positions, such as status hierarchies and group memberships, provides a conceptual and methodological framework for studying inequality and heterogeneity, which are the two fundamental concepts for social structure analysis. Third, the embeddedness of individuals in social relationships with their reference groups, such as family, friends, and acquaintances, offers insight into socialisation and social influence processes, within which individuals' perceptions and attitudes may change according to the composition of the social environments they inhabit. Consequently, the social distance generated by the network and spatial segregation isolated individuals from the lives of others, their experiences, and lifestyles. By contrast, more heterogeneous social networks imply broader and more accurate images of society that improve social bonding and solidarity. In sum, synthesising these three literature corpora provides the dissertation's theoretical framework, which will be empirically addressed in the following section.

\newpage 

# Research plan  

Considering the theoretical and empirical elements presented above, I would like to contribute to the redistributive preferences literature and discuss the role of social networks. As a result, my main research question is *What is the role of social networks on redistributive preferences?* and will be addressed in three independent studies. Figure  \@ref(fig:theory-model)  represents the theoretical framework of the research project.

```{r theory-model, echo = FALSE, message=FALSE, fig.align='center', fig.cap='Theoretical framework of the dissertation', out.width='1.5\\linewidth', fig.pos='H'}
knitr::include_graphics("images/theory-model.pdf")
```

## Study 1: "Economic inequality, social networks, and redistributive preferences: A cross-national study of 33 countries"

This study will focus on the relationship between social network segregation and redistributive preferences. On the one side, part of the literature suggests that social networks directly affect preferences for redistribution and welfare, arguing that access to diverse social positions through social ties explains changes in attitudes through social influence and socialisation mechanisms [@lindh_missing_2021;@paskov_crossclass_2022]. Therefore, it is expected that contact diversity in social networks increases support for redistribution (H1). On the other hand, part of the literature suggests that redistributive preferences are influenced by network diversity through perception, arguing that attributes of reference groups in social networks shape people's views about economic inequality in homogenous reference groups via subjective sampling mechanism [@KelleyEvans1995;@londono-velez_impact_2022] leading them to perceived society as more egalitarian and supporting less redistribution [@garcia-castro_perceived_2022]. Thus,  a second inquiry is to clarify if the influence of network diversity on redistributive preferences is mediated by perceived economic inequality (H2). Finally, comparative research on social capital has suggested that economic inequality strengthens the current spatial and social distance between the lower and upper classes in the social structure, decreasing their meeting opportunities as well as their chances of establishing more diverse social environments  [@otero_lives_2022a;@pichler_social_2009]. Consequently, it is expected that the association between network diversity and redistributive preferences to be stronger in more unequal societies (H3). 

It is expected to employ data from the "Social Networks and Social Resources" module from the International Social Survey Programme [@isspresearchgroup_international_2019], a large-scale nationally representative sample of 44,492 individuals in 33 countries. Regarding our main dependent variable, the survey includes a question regarding the extent that respondents agree that it is the government's responsibility to reduce income differences between high and low-income people, measured through a 1 to 5 Likert scale. Concerning the network structure, the survey includes the position generator that includes 10 occupations (e.g. Manager, school teacher, or car mechanic), enough for representing the diversity of positions in the occupational structure [@joye_measuring_2019]. For perceived inequality, the survey includes a general measure of perceived economic inequality ('Diff. in income are too large'), perceived household economic constraints ('How difficult or easy is it currently to your household to make ends meet (monthly)''), as well as the 10-rung MacArthur scale of subjective status ('top-bottom') 

Given the hierarchical structure of the data (individuals nested in countries), a multilevel regression model [@Hox2008] will be used for testing the association of network heterogeneity on redistributive preferences (H1). Consequently, for testing the indirect effect of network diversity via perceived inequality (H2), mediation analysis will be used [@imai_general_2010].  Finally, the moderation of economic inequality on the association of network diversity  (H3) will be modeled using cross-level interactions with random slopes [@aguinis_bestpractice_2013].

## Study 2 : "Residential segregation, social networks and market distribution of social services: longitudinal evidence"

The study seeks to understand the influence of social networks on support for the market provision of social services [@lindh_public_2015]. Consequently, it is suggested that network segregation increases the social distance between social classes and diminishes the sense of attachment to society  [@otero_lives_2022]. This lack of common experiences between people increases the social distance between groups and diminishes the chances of being integrated into social life, which in turn threatens social integration and cohesion [@sachweh_moral_2012]. Therefore, social solidarity can be eroded in societies where the sense of belonging weakens and the support for universalistic welfare schemes might not be strong among citizens, which leads to expect that network segregation fosters support for welfare marketisation (H1).  Additionally, societies that prosper economically in an unequal way imply higher network homogeneity which has a corollary in terms of urban residential segregation [@mendez_upper_2019]. Thus, the upper class can choose to develop their life plan in exclusive and isolated communities while the lower class is displaced to less advantaged neighborhoods and deprived from proper social services, which has consequences in the access to social resources [@otero_space_2022]. Consequently, living in more segregated communities will nurture isolated life experiences that are expected to increase support for welfare marketization (H2). Additionally, given that network homogeneity is higher in spatially segregated communities, it is expected that the influence of network segregation on welfare marketisation preferences is stronger when spatial segregation is high (H3).  

It is planned to use data from the Chilean Longitudinal Social Survey (ELSOC)  [@coes_estudio_2018]. First, on the substantive side, Chile has been recognised among Latin American countries for its economic and social development model since the 1980s, characterised as a prosperous society in terms of economic growth and also for introducing a strongly market-oriented welfare scheme [@arrizabalo_milagro_1995a]. In this context, during the 1990s, the expansion of tertiary education boosted social mobility, contributing to the size of the intermediate classes and the relevance of the service sector [@Torche2005a]. Additionally, the transformation of the class structure has not necessarily been associated with an equal distribution of salaries [@EspinozaNunez2014], considering that income distribution has remained relatively stable during the past decades [@Rodriguez2018]. Moreover, jointly with the transformation of the social structure, public concerns about social and economic inequality have increased in Chilean society [@pnud2017-desiguales], and the public debate on how to address the connection between redistribution and inequality through policy-making [see @atria_impuestos_2022]. 
Second, on the empirical side, ELSOC is an individual longitudinal panel survey representative of the urban areas of Chile (2016–2022, N = 2,927) that has repeatedly interviewed a sample of individuals on conflict and social cohesion-related topics, including social networks and economic inequality [^1]. Additionally, the sample has been designed to include a sufficient number of territorial units, as well as a sufficient number of cases within each one, to allow multilevel models grouped by city or municipality [@Snijders2011].

[^1]: For more details on the sampling and questionnaire design, visit: https://coes.cl/encuesta-panel/ 

The marketisation of welfare provision can be measured by employing three questions regarding the level of agreement with the statement that "people with higher income can have access to better [social services]" (e.g. education, health, or pensions). Also, ELSOC includes a version of the position generator with 16 occupations that allows one to describe the number of accessed social positions in an ego-centred network. Finally, regarding the structure of spatial segregation, recent work has established a synthetic measure of the spatial structure of urban areas by drawing georeferenced administrative data at the *Commune* [^2] level (see @otero_space_2022). Therefore, the available administration allows one to observe territorial units' variations and time trends regarding segregation.

[^2]: The Political Constitution divides state powers into hierarchical territorial units, in which the Executive Power is divided into Regions, Provinces, and Communes. Therefore, Communes are the lowest local-level units.  For more details at https://www.bcn.cl/siit/nuestropais/division.htm (in Spanish).

Regarding empirical strategy, it is crucial to consider that individuals have common variances with themselves when analyzing data over time. It is also important to recognize that individuals are nested within communes, and if this is not considered, they will be assumed to be independent. Therefore, a longitudinal multilevel approach is adopted to obtain a better estimation of standard errors, which includes a random term to define this variance (random effects) in the regression estimation (see @singer_applied_2009).

## Study 3: "Reference groups similarity and social solidarity: A vignette study."

The literature on reference groups has suggested that the characteristics of these groups influence perceptions and evaluations of economic inequality [@Evans1992]. Also, the tendency to prefer similar people has been described extensively in social network research [@mcpherson_birds_2001]. In this line, it has been demonstrated that both demographic and socioeconomic characteristics of individuals and their reference groups influence the evaluations of social policies such as pensions or taxes [@castillo2019;@sachweh_deserving_2023]. Furthermore, the literature on social cohesion has suggested that norms and social values play a crucial role in the support and legitimacy of welfare policies as an expression of social solidarity [@esping-andersen_three_1990], understood as the degree to which people bond and have a shared purpose, including how willing they are to collaborate and strive for shared objectives and the degree of trust and support they give to one another within their community [@green_regimes_2011]. It has been demonstrated that cross-class diversity influence support for egalitarian welfare policies [@lindh_missing_2021], but nuances in social solidarity appear when diversity is perceived as a threat [@castillo_social_2023]. Consequently, the inquiry related to what extent similar or diverse reference groups can impact one's attitudes toward social solidarity, including being willing to assist others even if it may come at a personal cost, as well as expectations for social reciprocity. Thus, the guiding question in this regard would be *what is the role of reference groups' similarity in attitudes toward social solidarity?*

A factorial vignette study will be implemented to address the questions presented above. This method allows the researcher to understand the role of the observed characteristic of a fictitious subject or vignette in the external observer evaluation  [@Jasso1990]. Therefore, the evaluations are adjusted according to the attributes of the vignettes [@dulmer_factorial_2016]. Also, each vignette is described in terms of specific demographic or socioeconomic characteristics (e.g. gender, occupation), which the survey participants consequently evaluate. For efficiency reasons, each respondent evaluates only a limited number of vignettes to guarantee between and within-subject in the expected responses  (e.g. ten vignettes ).  

Regarding the target population, it would be possible to implement the study in Chile, employing the current sample of the ELSOC survey in Chile (N $\approx$ 1500) [@elsoc_estudio_2018], as well as the online sample of the Cohesion Panel of the Research Institute for Social Cohesion (N $\approx$ 4000) in Germany [^3]. Also, given that the socioeconomic and demographic characteristics are already registered, it will be possible to match this data to the responses to the survey. It expected to implement the questionnaire using an online survey designed for this purpose. Therefore, the survey flow will include two main sections: (1) introduction and (2) vignette evaluation. Two sets of questions will be asked regarding the measures for social solidarity. First, *Concerning this person, you would be willing to:* (1) "Lend him/her money", (2) "Help him/her get a job", and (3) "Talk to him/her to offer emotional support". Second, *Now, you would expect this person to be willing to*: (1) "Lend you money", (2) "Help you get a job", and (3) "Talk to you to offer emotional support". The responses will be measured using an 11-point scale where 0 is "not willing at all", and 10 is "completely willing".

[^3]:For more details see: https://www.fgz-risc.de/forschung/datenzentrum 

The working plan of the dissertation is specified in Table \@ref(tab:workplan), considering the three independent studies for the following months.
```{r workplan, echo=FALSE}
pacman::p_load(knitr,kableExtra)
options(knitr.kable.NA = '')
xlsx::read.xlsx(file = "images/working_plan.xlsx",sheetIndex = 1) %>% 
  kable(col.names = c("Task", 1:10),
        booktabs=T,
        format = "latex",
        caption = "Working plan") %>%
  kable_styling(full_width = F,
   latex_options = c("striped","HOLD_position"),
   font_size = 7,
   position = "center") %>%
  add_header_above(c(" " = 1, "Month" = 10)) %>% 
  pack_rows("Study 1", 1, 5) %>%
  row_spec(5, hline_after = T) %>% 
  pack_rows("Study 2", 6, 10) %>%
  row_spec(10, hline_after = T) %>% 
  pack_rows("Study 3", 11, 18)
```





<!-- *How much would you like to become friends with her/him?*, from not at all (0) to very much (10)  [see @walther_attitudes_2023] -->

<!-- - gender (2)
- age (4)
- national/non-national (2)
- social class/occupations (5) - based on EGP5 
- education (4) - 
- income (5) -  quintiles

- Argumento reference groups homophily ?
- Argumento deservingness e inclusion on social welfare schemes 
- Rationale comparative study (north/south, social structure, institutional dimension)
- Vignette's dimensions (e.g. occupations, nationality)
- Match respondent-vignette ('homophily argument'). Does the match of certain characteristics influence deservigness? 

The third part of my research will incorporate elements from the distributive justice literature to provide a complete image of the role of reference groups in redistributive preferences. In this regard, evaluations are affected by observer's and observed characteristics, allowing us to estimate how these factors impact an outcome. 

Therefore, I propose a factorial vignette design to assess the role vignette's attribute on the observer evaluation of the hypothetical individuals being affected or included by specific social policies (e.g. increased taxes, health or unemployment benefits).  


Idea B: 

The study aims to investigate the role of reference groups in shaping perceptions of deservingness of social policies. 

The research design will involve a vignette study with a sample of participants from diverse backgrounds. 



The scenario will vary systematically in terms of the demographic characteristics of the person in need (e.g., age, gender, race/ethnicity) and the type of assistance needed. Participants will then be asked to rate the deservingness of the individual in the scenario to receive social assistance. 

We will also collect data on participants' own demographic characteristics, social status, and their reference groups in terms of social networks. We will examine the extent to which perceptions of deservingness are influenced by the demographic characteristics of the person in need, the type of assistance needed, and the participant's own demographic and social characteristics. 

Additionally, we will investigate whether the homophily of reference groups plays a role in shaping perceptions of deservingness, by examining whether participants' ratings of deservingness vary based on the demographic characteristics of their reference groups. 

The study aims to shed light on the ways in which reference groups and the homophily of reference groups may shape perceptions of deservingness and have implications for social policy. -->